---
title: "Shiny Application Tutorial"
author: "Rich Lavery"
date: "March 15, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Shiny Application Tutorial

All of the code for this project can be found in the following GitHub location:
`https://github.com/rjl2168/Shiny-Web-App-Tutorial/`

In this tutorial we will go over the basic features and workings of what makes up a Shiny application.  If you haven't already done so, be sure to install the "shiny" package using `install.packages("shiny")`.  This tutorial also makes use of the `ggplot2` library as well.

### UI Definition

```{r eval=FALSE}
library(shiny)
library(ggplot2)

# Define UI for application that draws a histogram
ui <- fluidPage(
   
   # Application title
   titlePanel("Interactive Shiny Application Example"),
   
   sidebarLayout(
```

We will start by loading the `shiny` library as well as any other libraries we will need for our application. 
Next we will start the definition of the User Interface or `ui`.  We define the `ui` using the `fluidPage()` function which will render an HTML file to display the components of our application.  Within the `fluidPage()` we define a title for the app using `titlePanel()`.  We can then begin to layout the page using the `sidebarLayout()` function.

#### Inputs

```{r eval=FALSE}
      #The sidebarPanel shows the various inputs
      sidebarPanel(
         selectInput(inputId = "u",
                     label = "Mean",
                     choices = c("-10", "-5", "0", "5", "10"),
                     selected = "0"),

         selectInput(inputId = "std",
                    label = "Standard Deviation",
                    choices = c("1", "3", "5"),
                    selected = "1"),
         
         radioButtons(inputId = "NumSamp",
                      label = "Number of Samples",
                      choices = c("100", "500", "1000"),
                      selected = "500"),

         sliderInput(inputId = "NumBins",
                     label = "Number of Bins in Histogram",
                     min = 1,
                     max = 50,
                     value = 30),
         
         actionButton(inputId = "showStats",
                      label = "Print Stats")),
```

In our example, the first element of the `sidebarLayout()` is the `sidebarPanel()` which will contain all of our user inputs.  Here we have a total of five user inputs demonstrating four different input types including (1) two selection boxs, (2) a list of radio buttons, (3) a slider input, and (4) an action button.  There are other inputs types as well.  You, as the developer, can add as many user inputs as necessary.  Each user input will have an `inputId` and a `label`.  The `inputId` is needed for referencing the object later in the server application code, whereas the `label` is simply a discriptive text string the application user will see when running the app.  In addition to these first two attributes there are additional attributes that may be specific to that particular input function.

In our example we are allowing the user to select one of five different mean values and one of three different standard deviations using the `selectInput()` function which is displayed as a list of choices in a pull-down selection box.

Next, we give the user the option to select one of three different sample sizes for our normal sample set.  This is done with the `radioButtons()` function which is displayed as a list of three options with a radio button for each choice.

Next, we give the user the ability to select the number of bins in our histogram plot using the `sliderInput()` function which is displayed as a slider than ranges between the `min` and `max` values defined.

Finally, we provide the user with a button using the `actionButton()` function which will display the statistics for our normal sample (mean and standard deviation) when the user pushes the button.

#### Outputs

```{r eval=FALSE}
      # The mainPanel calls out placeholders
      mainPanel(
         plotOutput("distPlot"),
         textOutput("meanValue"),
         textOutput("stdValue")
      )
   )
)
```

The next element of the `sidebarLayout()` is the `mainPanel()` which will generate place holders in the HTML page for all of the outputs we intend to render.  In this case we will need a space to plot the normal sample distribution which is denoted by the `plotOutput("distPlot")` line.  We also have provisions for two text outputs to show the mean and standard deviation of the sample set.  This is called out with the `textOutput("meanValue")` and `textOutput("stdValue")` lines.  Similar to the `id` string on the input functions, the text string inside the output function argument will be used to link the rendered outputs that are generated in the server code below.

### Server Definition

```{r eval=FALSE}
# Define server logic required to draw a histogram
server <- function(input, output) {
   #Create a reactive object `sample_data` which can be used to other render functions
   sample_data <- reactive({
     Sample_size <- as.numeric(input$NumSamp)
     Avg <- as.numeric(input$u)
     Std_dev <- as.numeric(input$std)
     rnorm(n = Sample_size, mean = Avg, sd = Std_dev)
   })
```

We now move onto the second part of the shiny application which is the server function code.  The server function works off the input and output elements that were defined in the user interface portion of the application code.  

We first create a reactive object called `sample_data` using the `reactive()` function.  We pass into the function a block of R code contained in the `{}` brackets.  Notice we are pulling in values for the sample size, mean value, and standard deviation value from the user input functions defined in the `ui`.  We do this by calling `input$<id>` where the <`id`> is the `id` attribute of our input.  Notice there are no `""` surrounding the `id`.  We then use these three values to generate the random normal sample using `rnorm`.  This sample list then gets returned to our `sample_data()` object for future use in the app.

One important note on "reactivity".  Whenever the user changes one or more of the three reactive inputs that go into calculating the `sample_data`, the server will re-run this whole block of code resulting in new `sample_data` values.

```{r eval=FALSE}
   #Create a renderPlot object to update the plot when either the `sample_data` changes or the `NumBins` changes
   output$distPlot <- renderPlot({
     Num_bins <- input$NumBins
     samples_df <- as.data.frame(sample_data())
     ggplot(samples_df, aes(x = samples_df$sample_data)) + geom_histogram(aes(y = ..density..), 
                                                                          stat = "bin", bins = Num_bins) +  
       geom_density() + coord_cartesian(xlim = c(-20, 20)) + xlab("Data Values")
   })
```

The next part of our server code is to render a plot object using the `renderPlot()` function that will be linked to the plot area we defined in the `ui` code using `output$disp`.

Notice that there are two reactive inputs that this plot relies on: (1) `input$NumBins` from the slider input and (2) `sample_data()` which is the reactive value we calculated above.  If either of those items change, the server will re-run the code in the `renderPlot()` function to regenerate the plot.

```{r eval=FALSE}
   #Create a reactive output that only updates upon observing an event, such as the user pressing a button
   observeEvent(input$showStats, {
     output$meanValue <- renderText(isolate({paste("Mean =", toString(mean(sample_data())), sep = " ")}))
     output$stdValue <- renderText(isolate({paste("Standard Deviation =", toString(sd(sample_data())), sep = " ")}))
   })
}
```

The last element of our server code is a function called `observeEvent()`.  We use this function to trigger a block of code to run when a particular event occurs.  In this case, that event is the user pressing the action button we defined in the `ui`.  When the action button is pressed, we will run two `renderText()` functions and link them to the two text outputs `output$meanValue` and `output$stdValue`.  You'll also notice we are using an `isolate()` function to encapsulate the R code.  We do this because the code contains the reactive element `sample_data()` and we do not want the code to run when `sample_data()` changes.  We only want the code to run when the user presses the action button.

### Build the Shiny Application

```{r eval=FALSE}
# Bind the user interface definition with the server definition and run the application 
shinyApp(ui = ui, server = server)
```

The third and final component of any shiny application is the `shinyApp()` function which binds the `server` code to the `ui` code and builds the app.

## Try out the app: [https://rjl2168.shinyapps.io/shiny-web-app-tutorial/](https://rjl2168.shinyapps.io/shiny-web-app-tutorial/)
